---
title: "Creating MAMS from AnnData"
date: "`r Sys.Date()`"
package: rmams
output:
    rmdformats::readthedown:
      toc_depth: 4
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```



# Introduction

Matrix and Analysis Metadata Standards (MAMS) can be used to capture the relevant information about the data matrices and annotations that are produced during common and complex analysis workflows for single-cell data.

This workflow highlights how an RNA expression based experiment from an AnnData object can be annotated using the rmams package. For this purpose, we will analyze the [PBMC 1K dataset](https://support.10xgenomics.com/single-cell-gene-expression/datasets/4.0.0/SC3_v3_NextGem_SI_PBMC_CSP_1K) using Python based packages and AnnData as our data-container.

The steps include:
1. Importing the raw/droplet and filtered/cell matrices
2. Performing quality control and further subsetting the filtered matrix to only include high-quality cells
3. Clustering all cells
4. Clustering T-cells only

This workflow demonstrates how different subsets of cells/observations are produced due to QC/filtering or can be made based on different biological subsets. Note that the goal of this workflow is less about producing the best clustering results and cell type characterization. Rather it is focused on producing data/results that is commonly generated during real-world analyses which can be annotated with MAMS. For more information about MAMS, see the [GitHub repository](https://github.com/single-cell-mams/mams)

## Step 1: Creating the SingleCellExperiment object

### Importing 

```{r library, message=FALSE}
library(reticulate)
library(anndata)
sc <- reticulate::import("scanpy")
```

First, we will import raw/droplet and filtered/cell matrices:

```{r import, warning=FALSE}
adata.raw <- sc$read_10x_mtx("../data/raw_feature_bc_matrix/", var_names = "gene_ids")
adata.filtered <- sc$read_10x_mtx("../data/filtered_feature_bc_matrix/", var_names = "gene_ids")

# ensembl_ids <- rownames(sce.filtered)
# gene_symbols <- mapIds(org.Hs.eg.db, keys = ensembl_ids, column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")
# rownames(sce.filtered) <- ifelse(is.na(gene_symbols), ensembl_ids, gene_symbols)
```

### Quality control

The filtered matrix will be further subsetted based on mitochondrial percentage:

```{r qc}
var_df <- adata.filtered$var
var_df$mito <- grepl("MT-", var_df$gene_symbols)
adata.filtered$var <- var_df
sc$pp$calculate_qc_metrics(adata.filtered,
                           qc_vars = list("mito"),
                           inplace = TRUE)
adata.filtered.nonartifact = adata.filtered[adata.filtered$obs$pct_counts_mito < 5]
```

### Clustering

The goal of many single cell workflows are to perform clustering to identify biologically distrinct groups of cells. We will perform clustering on all QC'ed cells using RNA, ADT, or both and then also re-cluster the T-cells by themselves.

### Analyzing all cells

#### RNA only clustering of all cells
```{r cluster_rna, message = FALSE, warning=FALSE}
sc$pp$log1p(adata.filtered.nonartifact)
sc$pp$highly_variable_genes(adata.filtered.nonartifact, min_mean = 0.0125, max_mean = 3, min_disp = 0.5)
sc$pp$scale(adata.filtered.nonartifact, max_value = 10)
sc$tl$pca(adata.filtered.nonartifact, svd_solver = "arpack")
sc$pp$neighbors(adata.filtered.nonartifact, n_neighbors = 10L, n_pcs = 40L)
sc$tl$umap(adata.filtered.nonartifact)
sc$tl$leiden(adata.filtered.nonartifact)
```

```{r plot_rna,message = FALSE, warning=FALSE}
sc$pl$umap(adata.filtered.nonartifact, color = list("leiden"))
```


```{r save_data,message = FALSE, warning=FALSE}
adata$write_h5ad(filename = "../inst/extdata/pbmc1k_adata_raw.h5ad")
adata.filtered$write_h5ad(filename = "../inst/extdata/pbmc1k_adata_filtered.h5ad")
adata.filtered.nonartifact$write_h5ad(filename = "../inst/extdata/pbmc1k_adata_filtered_nonartifact.h5ad")
```

## Step 2: Creating the MAMS object from the SingleCellExperiment object

### Loading data

```{r create_mams_object,message = FALSE, warning=FALSE}
library(rmams)

# get the file paths of anndata objects

file_paths <- c("../inst/extdata/pbmc1k_adata_raw.h5ad", "../inst/extdata/pbmc1k_adata_filtered.h5ad", "../inst/extdata/pbmc1k_adata_filtered_nonartifact.h5ad")

# get the names of these objects 
file_names <-  gsub(pattern = "\\.rds$", replacement = "", x = basename(file_paths))

# save the anndata objects onto the object_list by reading the object from file path.
object_list <- lapply(file_paths, read_h5ad)

# name the list objects using previously generated file_names.
names(object_list) <- file_names

# define the type of each subset
observational_subsets<-c("raw", "filtered","nonartifact")
datasetid = "PBMC1K"
```


### Create MAMS object

```{r create_object}
result<- convert_AnnData_to_MAMS(object_list,observational_subsets,datasetid)
```

### Accessing data 

```{r show_results}
# Get the dataset_id from the first fom
fom(result,"fom1","dataset_id")
```

## Step 3: Validating MAMS object

```{r validate_mams,message=TRUE,warning=TRUE, error=TRUE}
fom(result,"fom1","dataset_id") <-""
check_MAMS(mams_object = result)
```

### Step 4: Updating fields manually

```{r}
# add dataset id
fom(result,"fom1","dataset_id") <-"PBMC1K"

# by default all reduced dimensions are annotated as Reductions, but UMAP can be better reflected as Embedding
fom(result, "fom6", "processing") <- "Embedding"
```

we can add some commands as well
```{r}
# command example 
# show two examples, annotate counts matrix and annotate one of reducedDims, commands, params 
record_id <- "CellRanger.count"
record_package_name<- "CellRanger"
record_function_name<- "count"
record_package_version<- "unknown"

results@REC[[record_id]]<- create_REC_object(record_package_name = record_package_name,
                                          record_function_name = record_function_name,
                                          record_package_version = record_package_version)

```

## Storing mams back to object
```{r}
# uns slot
```


## Step 6: Saving MAMS object to JSON
### Here we show how you can convert the MAMS object to the JSON format for export. MAMS and JSON formats are interchangeable.

```{r}
write_MAMS_to_output(MAMS=result, filepath = "../inst/extdata/pbmc1k_rna_adata_mams.JSON",format = "JSON")
```

## Step 5: Saving MAMS as YAML

```{r}
write_MAMS_to_output(MAMS=result, filepath = "../inst/extdata/pbmc1k_rna_adata_mams.yml",format = "YAML")
```
